Ferromagnetism and spin excitations in topological Hubbard models with a flatband

### 2018-11-08 paper/paper.tex
For Submission to PRB.

\begin{document}
\title{Ferromagnetism and spin excitations in topological Hubbard models with a flatband}
\author{Xiao-Fei Su}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{School of Physics and Electronic Information, Huaibei Normal University, Huaibei 235000, China}
\author{Zhao-Long Gu}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\author{Zhao-Yang Dong}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\author{Shun-Li Yu}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{Collaborative Innovation Center of Advanced Microstructures, Nanjing University, Nanjing 210093, China}
\author{Jian-Xin Li}
\email[]{jxli@nju.edu.cn}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{Collaborative Innovation Center of Advanced Microstructures, Nanjing University, Nanjing 210093, China}
\date{\today}

@@

\par Another involved intriguing phenomenon arising from Coulomb repulsions in flat or nearly-flat bands is the itinerant ferromagnetism \cite{T_PTP1998}. It was proved by Tasaki and Mielke that the ground state of a flat electron band with a filling factor not more than but sufficiently close to $1/2$ is ferromagnetically ordered as long as an infinitesimal onsite Hubbard repulsion is present \cite{T_PRL1992,M_PLA1993,MT_CMP1993}. Afterwards, this ferromagnetism was shown to be stable against small nonflatness of the electron bands if and only if the Hubbard interaction exceeds a critical value \cite{T_PRL1994}. Spin wave excitations over this ferromagnetic ground state were also studied \cite{KA_PRL1994,SGDL_PRB2018} and itinerant topological magnons have been reported quite recently \cite{SGDL_PRB2018}.

@@

\subsection{Exact diagonalization with projection}\label{exact_diag_proj}
\par Exact diagonalization method with a projection onto the low-energy Hilbert space has been widely applied to systems that host flat or nearly-flat energy bands \cite{NSCM_PRL2011,RB_PRX2011,NSRCM_PRB2011,SGDL_PRB2018,NSRCM_PRL2012,DG_PRB2015}. This approach applies when the energy gap between the flat or nearly-flat band and other bands is larger than the Coulomb interaction.

@@

\par Before the detailed discussion on the spin-1 excitation spectra of the quarter-filled Chern Hubbard model and $Z_2$ Hubbard model, we present their phase diagrams first, which are shown in Fig. \ref{phase}. Here, the colormap represents the ratio of the Hubbard interaction strength $U$ to the lower electron bandwidth $W$. FM denotes the ferromagnetic phase and NFM denotes the non-ferromagnetic phase. As discussed in Sec. \ref{exact_diag_proj}, the phase boundary is determined as the onset point of $U$ below which negative spin-1 excitation spectra arise. The phase diagrams of these two models appear quite similar. It is obvious that a critical Hubbard interaction strength is needed to maintain the ferromagnetically ordered ground state when the electron band has finite nonflatness. Furthermore, the phase boundaries always lie near the contour line with $U/W=2.0$. Qualitatively speaking, these observations are expected as a result from the competition between the kinetic energy and potential energy of the electrons, as a fully spin-polarized state minimizes the energy of Hubbard interactions but cost more energy when the electron band disperses. In the following, we will elaborate the spin-1 excitation spectra of these models.

@@

\par In the flatband limit, the spin-1 excitation spectra as shown in Fig. \ref{fmcispectrum}(a) as the green dashed lines show several similarities with the dispersive case, including the constitution of the spin waves and Stoner continuum exhibiting as a flat line at $E=0.5U$, the gapless acoustic spin wave and its quadratical dispersion  near the $\Gamma$ point, and the degeneracy between the acoustic and optical branches along the $X$-$M$ path. Comparing results in the flatband limit with those obtained by the generalized bosonization method at the harmonic approximation [see Fig. 4 in Ref. \cite{DG_PRB2015}], one will find that two results are qualitatively quite consistent. In the scheme adopted in Ref. \cite{DG_PRB2015}, the ignorance of the interactions between magnons suggests that only the single-magnon excitations are captured. So, this consistence indicates that the interactions between magnons shows a negligible effect in the flatband limit. However, when the free band becomes dispersive, considerable distinct features appears. Now, the spin-wave energies around the $X=(0,\pi)$ and $O=(\pi/2,\pi/2)$ points are strongly suppressed, as a result two local minima are formed at $X$ and $O$ which can be ascribed to the emergence of the roton-like spin-wave excitations. Furthermore, the continuum exhibiting as a straight line in the flatband limit extends to a large grey region shown in Fig. \ref{fmcispectrum}(a) with dispersive boundaries, and the local minima of its lower boundary coincide with those in the spin-wave dispersion, as indicated by the black arrows in Fig. \ref{fmcispectrum}(a).

@@

\par This observation shows that it is the extending of the Stoner continuum arising from the nonflatness of the $d$ band that pushes down the spin-wave excitations. In particular, it suggests that the couplings between the individual excitations and spin waves would be responsible for the emergence of the roton-like spin waves.

@@

\par The emerged roton-like spin waves are essential for the destabilization of the ferromagnetic phase. In Fig. \ref{pbcispectrum}($\text{a}_1$) and \ref{pbcispectrum}($\text{b}_1$), we plot the spin-1 excitation spectra of the model near the FM/NFM phase boundary, with two sets of parameters marked by the blue stars in Fig. \ref{phase}(a). The corresponding illustrations of the band-splitting picture are shown in Fig. \ref{pbcispectrum}($\text{a}_3$) and \ref{pbcispectrum}($\text{b}_3$). Combined with Fig. \ref{fmcispectrum}(a), it can be seen that the increase of the nonflatness leads the boundary bottom marked by $q_1$ to move toward low energies further. In this process, a roton-like local minimum in the acoustic band appears nearby, which can be seen more clearly in Fig. \ref{pbcispectrum}($\text{a}_2$) and \ref{pbcispectrum}($\text{b}_2$). With the increase of the nonflatness of the electron band, the energy of this newly formed roton-like mode goes down and touches zero, which leads to the destabilization of the ground state. From Fig. \ref{pbcispectrum}($\text{a}_3$) and \ref{pbcispectrum}($\text{b}_3$), we can see that the ferromagnetic ground state is quite robust against the spin-1 flips until the top of the spin-up band is approaching near to the bottom of the spin-down band. Considering that the energy splitting of these two bands is $U/2$, the destabilization point would be quite close to $U/2\sim W$.

\subsection{$Z_2$ Hubbard model}\label{ZHM}
\begin{figure}
\includegraphics[width=\columnwidth]{tispectrum}
\caption{(Color online) (a1) and (b1): Spin-1 excitation spectra of the $Z_2$ Hubbard model. Green solid lines denote the spin waves and grey regions denote the Stoner continuum. Green dashed lines represent the corresponding spectra in the flatband limit. Black arrows mark the minima of the Stoner continuum. (a$_2$) and (b$_2$): Illustration of the splitting of the lower electron bands of the $Z_2$ Hubbard model. (a$_3$) and (b$_3$): Positions of the corresponding local maxima (red solid circles) of the spin-up bands and local minima (purple solid circles) of the spin-down bands in the first Brillouin zone. Green arrows denote the corresponding scattering channels of the minima of the Stoner continuum marked in (a$_1$) and (b$_1$). $t_2=0.650$ for (a) and $t_2=0.741$ for (b). Other parameters are fixed at $t_1=1.0$, $U=2.0$.}
\label{tispectrum}
\end{figure}

@@

\par Similar to the Chern Hubbard model, the introduction of the dispersion to the flatband leads to the appearance of three local minima (roton-like modes) along the $\Gamma-X-M-\Gamma$ direction for both acoustic and optical bands, respectively. The momentum positions of these minima coincide the corresponding boundary bottoms of the Stoner continuum, suggesting that the renormalization of the spin-wave bands is due to the coupling between the spin waves and Stoner continuum. In the band-splitting picture as proposed above, the position of a local boundary bottom of the Stoner continuum is determined by the transfer momentum of the particle-hole scattering between a spin-up maximum and a spin-down minimum. Therefore, we present in Fig. \ref{tispectrum}($\text{a}_2$) and \ref{tispectrum}($\text{b}_2$) the dispersions for the lower electron bands in the $Z_2$ Hubbard model showing the splitting of the spin-up and spin-down bands. The corresponding local maxima (red solid circles) of the spin-up bands and local minima (purple solid circles) of the spin-down bands in the first Brillouin zone are shown in Fig. \ref{tispectrum}($\text{a}_3$) and \ref{tispectrum}($\text{b}_3$). One can see that the momentum positions of the boundary bottoms as marked by the black arrows in Fig. \ref{tispectrum}($\text{a}_1$) and Fig. \ref{tispectrum}($\text{b}_1$) (here only $q_1$ is plotted for illustration) match the transferred momenta between the spin-up and spin-down bands, which gives a firm support of the picture based on the band splitting.

@@

\par We would like to remark that the physics of the downward renormalization of the energies of the collective modes affected by the boundary bottoms of the Stoner continuum shares similarity with that observed in the two-dimensional antiferromagents, where the spin waves are interpreted as bound states of confined spinons and the suppression of energies of spin waves around the continuum bottom is attributed to their couplings to the nearby deconfined spinon continuum\cite{ZFSMC_PRL2006,TS_PRL2013,DMCNTPEMIR_NP2015,SQCCMS_PRX2017,YWDYL_PRB2018}. In addition, although correlated nearly-flat topological bands are hard to realize in real materials, they are possible to be designed in cold-atom systems. In fact great progress has been made very recently \cite{LCJPS_N2009,AALBPB_PRL2013,MSKBK_PRL2013,JMDLUGE_N2014}. We expect our results will stimulate related investigations in future cold-atom experiments.

\begin{acknowledgments}
\par This work was supported by the National Natural Science Foundation of China (11774152) and National Key Projects for Research and Development of China (Grant No. 2016YFA0300401).
\par X.-F. S. and Z.-L. G. contributed equally to this work.
\end{acknowledgments}

>>>>

\begin{document}
\title{Ferromagnetism and spin excitations in topological Hubbard models with a flatband}
\author{Xiao-Fei Su}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{School of Physics and Electronic Information, Huaibei Normal University, Huaibei 235000, China}
\author{Zhao-Long Gu}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\author{Zhao-Yang Dong}
\affiliation{Department of Applied Physics, Nanjing University of Science and Technology, Nanjing 210094, China.}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\author{Shun-Li Yu}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{Collaborative Innovation Center of Advanced Microstructures, Nanjing University, Nanjing 210093, China}
\author{Jian-Xin Li}
\email[]{jxli@nju.edu.cn}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{Collaborative Innovation Center of Advanced Microstructures, Nanjing University, Nanjing 210093, China}
\date{\today}

@@

\par Another involved intriguing phenomenon arising from Coulomb repulsions in flat or nearly-flat bands is the itinerant ferromagnetism \cite{T_PTP1998,T_PRL1992,M_PLA1993,MT_CMP1993,KMTT_EPL2010}. It was proved by Tasaki and Mielke that the ground state of a flat electron band with a filling factor not more than but sufficiently close to $1/2$ is ferromagnetically ordered as long as an infinitesimal onsite Hubbard repulsion is present \cite{T_PRL1992,M_PLA1993,MT_CMP1993}. Afterwards, this ferromagnetism was shown to be stable against small nonflatness of the electron bands if and only if the Hubbard interaction exceeds a critical value \cite{T_PRL1994}. Spin wave excitations over this ferromagnetic ground state were also studied \cite{KA_PRL1994,SGDL_PRB2018} and itinerant topological magnons have been reported quite recently \cite{SGDL_PRB2018}.

@@

\subsection{Exact diagonalization with projection}\label{exact_diag_proj}
\par Exact diagonalization method with a projection onto the low-energy Hilbert space has been widely applied to systems that host flat or nearly-flat energy bands \cite{NSCM_PRL2011,RB_PRX2011,NSRCM_PRB2011,SGDL_PRB2018,NSRCM_PRL2012}. This approach applies when the energy gap between the flat or nearly-flat band and other bands is larger than the Coulomb interaction.

@@

\par Before the detailed discussion on the spin-1 excitation spectra of the quarter-filled Chern Hubbard model and $Z_2$ Hubbard model, we present their phase diagrams first, which are shown in Fig. \ref{phase}. Here, the colormap represents the ratio of the Hubbard interaction strength $U$ to the lower electron bandwidth $W$. FM denotes the ferromagnetic phase and NFM denotes the non-ferromagnetic phase. As discussed in Sec. \ref{exact_diag_proj}, the phase boundary is determined by the onset parameter at which the spin-1 excitation spectrum starts to acquire zero energy at a finite momentum. The phase diagrams of these two models appear quite similar. It is obvious that a critical Hubbard interaction strength is needed to maintain the ferromagnetically ordered ground state when the electron band has finite nonflatness. Furthermore, the phase boundaries always lie near the contour line with $U/W=2.0$. Qualitatively speaking, these observations are expected as a result from the competition between the kinetic energy and potential energy of the electrons, as a fully spin-polarized state minimizes the energy of Hubbard interactions but cost more energy when the electron band disperses. In the following, we will elaborate the spin-1 excitation spectra of these models.

@@

\par In the flatband limit, the spin-1 excitation spectra as shown in Fig. \ref{fmcispectrum}(a) as the green dashed lines show several similarities with the dispersive case, including the constitution of the spin waves and Stoner continuum exhibiting as a flat line at $E=0.5U$, the gapless acoustic spin wave and its quadratical dispersion  near the $\Gamma$ point, and the degeneracy between the acoustic and optical branches along the $X$-$M$ path. Comparing results in the flatband limit with those obtained by the generalized bosonization method at the harmonic approximation [see Fig. 4 in Ref. \cite{DG_PRB2015}], one will find that two results are qualitatively quite consistent. In the scheme adopted in Ref. \cite{DG_PRB2015}, the ignorance of the interactions between magnons suggests that only the single-magnon excitations are captured. So, this consistence indicates that the interactions between magnons shows a negligible effect in the flatband limit. However, when the free band becomes dispersive, considerable distinct features appears. Now, the spin-wave energies around the $X=(0,\pi)$ and $O=(\pi/2,\pi/2)$ points are strongly suppressed, as a result two local minima are formed at $X$ and $O$ which can be ascribed to the emergence of the roton-like spin-wave excitations. Furthermore, the continuum exhibiting as a straight line in the flatband limit extends to a large grey region shown in Fig. \ref{fmcispectrum}(a) with dispersive boundaries, and the local minima of its lower boundary coincide with those in the spin-wave dispersion, as indicated by the black arrows in Fig. \ref{fmcispectrum}(a). This observation shows that it is the extending of the Stoner continuum arising from the nonflatness of the $d$ band that pushes down the spin-wave excitations. In particular, it suggests that the couplings between the individual excitations and spin waves would be responsible for the emergence of the roton-like spin waves.

@@

@@

\par The emerged roton-like spin waves are essential for the destabilization of the ferromagnetic phase. In Fig. \ref{pbcispectrum}($\text{a}_1$) and \ref{pbcispectrum}($\text{b}_1$), we plot the spin-1 excitation spectra of the model near the FM/NFM phase boundary, with two sets of parameters marked by the blue stars in Fig. \ref{phase}(a). The corresponding illustrations of the band-splitting picture are shown in Fig. \ref{pbcispectrum}($\text{a}_3$) and \ref{pbcispectrum}($\text{b}_3$). Compared with Fig. \ref{fmcispectrum}(a), it can be seen that the increase of the nonflatness leads the boundary bottom marked by $q_1$ to move toward low energies further. In this process, a roton-like local minimum in the acoustic band appears nearby, which can be seen more clearly in Fig. \ref{pbcispectrum}($\text{a}_2$) and \ref{pbcispectrum}($\text{b}_2$). With the increase of the nonflatness of the electron band, the energy of this newly formed roton-like mode goes down and touches zero, which leads to the destabilization of the ground state. From Fig. \ref{pbcispectrum}($\text{a}_3$) and \ref{pbcispectrum}($\text{b}_3$), we can see that the ferromagnetic ground state is quite robust against the spin-1 flips until the top of the spin-up band is approaching near to the bottom of the spin-down band. Considering that the energy splitting of these two bands is $U/2$, the destabilization point would be quite close to $U/2\sim W$.

\subsection{$Z_2$ Hubbard model}\label{ZHM}
\begin{figure}
\includegraphics[width=\columnwidth]{tispectrum}
\caption{(Color online) (a$_1$) and (b$_1$): Spin-1 excitation spectra of the $Z_2$ Hubbard model. Green solid lines denote the spin waves and grey regions denote the Stoner continuum. Green dashed lines represent the corresponding spectra in the flatband limit. Black arrows mark the minima of the Stoner continuum. (a$_2$) and (b$_2$): Illustration of the splitting of the lower electron bands of the $Z_2$ Hubbard model. (a$_3$) and (b$_3$): Positions of the corresponding local maxima (red solid circles) of the spin-up bands and local minima (purple solid circles) of the spin-down bands in the first Brillouin zone. Green arrows denote the corresponding scattering channels of the minima of the Stoner continuum marked in (a$_1$) and (b$_1$). $t_2=0.650$ for (a) and $t_2=0.741$ for (b). Other parameters are fixed at $t_1=1.0$, $U=2.0$.}
\label{tispectrum}
\end{figure}

@@

\par Similar to the Chern Hubbard model, the introduction of the dispersion to the flatband leads to the appearance of three local minima (roton-like modes) along the $\Gamma-X-M-\Gamma$ direction for both acoustic and optical bands, respectively. The momentum positions of these minima coincide the corresponding boundary bottoms of the Stoner continuum, suggesting that the renormalization of the spin-wave bands is due to the coupling between the spin waves and Stoner continuum. In the band-splitting picture as proposed above, the position of a local boundary bottom of the Stoner continuum is determined by the transfer momentum of the particle-hole scattering between a spin-up band maximum and a spin-down band minimum. Therefore, we present in Fig. \ref{tispectrum}($\text{a}_2$) and \ref{tispectrum}($\text{b}_2$) the dispersions for the lower electron bands in the $Z_2$ Hubbard model showing the splitting of the spin-up and spin-down bands. The corresponding local maxima (red solid circles) of the spin-up bands and local minima (purple solid circles) of the spin-down bands in the first Brillouin zone are shown in Fig. \ref{tispectrum}($\text{a}_3$) and \ref{tispectrum}($\text{b}_3$). One can see that the momentum positions of the boundary bottoms as marked by the black arrows in Fig. \ref{tispectrum}($\text{a}_1$) and Fig. \ref{tispectrum}($\text{b}_1$) (here only $q_1$ is plotted for illustration) match the transferred momenta between the spin-up and spin-down bands, which gives a strong support of the picture based on the band splitting.

@@

\par We would like to remark that the downward renormalization of the energies of the collective modes affected by the boundary bottoms of the Stoner continuum shares similarity with that observed in the two-dimensional antiferromagents, where the spin waves are interpreted as bound states of confined spinons and the suppression of energies of spin waves around the continuum bottom is attributed to their couplings to the nearby deconfined spinon continuum\cite{ZFSMC_PRL2006,TS_PRL2013,DMCNTPEMIR_NP2015,SQCCMS_PRX2017,YWDYL_PRB2018}. In addition, although correlated nearly-flat topological bands are hard to realize in real materials, they are possible to be designed in cold-atom systems. In fact great progress has been made very recently \cite{LCJPS_N2009,AALBPB_PRL2013,MSKBK_PRL2013,JMDLUGE_N2014}. We expect our results will stimulate related investigations in future cold-atom experiments.

\begin{acknowledgments}
\par This work was supported by the National Natural Science Foundation of China (No.11774152, No.11674158) and National Key Projects for Research and Development of China (Grant No. 2016YFA0300401).
\par X.-F. S. and Z.-L. G. contributed equally to this work.
\end{acknowledgments}
